Active learning for screening prioritization in systematic reviews

### 2020-07-29 manuscript/manuscript/Ferdinands, G - MSBBSS.Rmd
update 29-07 corrections

---
title: Active learning for efficient systematic reviews
subtitle: Evaluating models across research areas 
author: "Gerbrich Ferdinands"
# keywords:
header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage[left]{lineno}
    - \usepackage{amssymb}
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  bookdown::pdf_document2:
    extra_dependencies: subfig, algorithm2e, float
    number_sections: no
    toc: no
  word_document: default
bibliography: asreview.bib
link-citations: yes
csl: elsevier-vancouver.csl
---

<!-- --- -->
<!-- title: Manuscript drafts -->
<!-- author: "Gerbrich Ferdinands" -->
<!-- date: '`r format(Sys.time(), "%d %B, %Y")`' -->
<!-- output: -->

<!--   bookdown::pdf_document2: -->
<!--     extra_dependencies: subfig, algorithm2e, float -->
<!--     number_sections: no -->
<!--     toc: no -->
<!--   word_document: default -->
<!-- bibliography: asreview.bib -->
<!-- link-citations: yes -->
<!-- csl: systematic-reviews.csl -->

@@

```{r all tables, results = 'hide'}
table_nums <- captioner(prefix = "Table")
# table  on datasets statistics
table_nums(name = "datasets", caption = "Statistics on the datasets obtained from six original systematic reviews.")
table_nums(name = "atd", caption = "ATD values ($\\bar x (\\hat s)$) for all model-dataset combinations, and median (MAD) for all datasets.")
table_nums(name = "wss", caption = "WSS@95 values ($\\bar x (\\hat s)$) for all model-dataset combinations, and median (MAD) for all datasets.")
table_nums(name = "rrf", caption = "RRF@10 values ($\\bar x, (\\hat s)$) for all model-dataset combinations, and median (MAD) for all datasets.")

@@

@@

\linenumbers
# Abstract
#### Background 
the context and purpose of the study

#### Methods
how the study was performed and statistical tests used

#### Results
the main findings

#### Conclusions
brief summary and potential implications

keywords:

# Background
Systematic reviews are top of the bill in research. A systematic review brings together all studies relevant to answer a specific research question [@PRISMA-PGroup2015]. Systematic reviews inform practice and policy [@Gough2002] and are key in developing clinical guidelines [@Chalmers2007]. However, systematic reviews are costly because they involve the manual screening of thousands of titles and abstracts to identify publications relevant to answering the research question.

<!-- An experienced reviewer takes on average 30 seconds to screen one title and abstract, whereas an inexperienced reviewer takes even longer [@Wallace2010]. A typical review of 5000 publications would require over 40 hours of uninterrupted screening, while typically only a fraction of the publications is relevant (mean = 2.94%, IQR=2.5 @Borah2017]). -->
Conducting a systematic review typically requires over a year of work by a team of researchers [@Borah2017]. Nevertheless, systematic reviewers are often bound to a limited budget and timeframe. Currently, the demand for systematic reviews exceeds the available time and resources by far [@Lau2019]. Especially when the need for guidelines is urgent it is extremely challenging to provide a review that is both timely and comprehensive.

To ensure a timely review, reducing workload in systematic reviews is essential. With advances in Machine Learning (ML), there has been wide interest in tools to reduce workload in systematic reviews [@Harrison2020]. Various learning models have been proposed, aiming to predict whether a given publication is relevant or irrelevant to the systematic review. Previous findings suggest that such models potentially reduce workload with 30-70% at the cost of losing 5% of relevant publications, i.e. 95% recall [@OMara-Eves2015].
<!-- To date, several studies have begun to examine the use of learning algorithms for screening prioritization. -->

A well-established approach in increasing efficiency in title and abstract screening is screening prioritization [@Cohen2009; @Shemilt2014]. In screening prioritization, the learning model presents the reviewer with the publications that are most likely to be relevant first, thereby expediting the process of finding all of the relevant publications. Such an approach allows for substantial time-savings in the screening process as the reviewer can decide to stop screening after a sufficient number of relevant publications have been retrieved [@Yu2019]. Moreover, the early retrieval of relevant publications facilitates a faster transition of those publications to the next steps in the review process [@Cohen2009]. 
<!-- Additionally, several studies report increasing efficiency beyond saving time [@OMara-Eves2015].  -->

<!-- assist the reviewer in the screening process -->
<!-- Several studies have proposed active learning models for screening prioritization -->
Recent studies have demonstrated the effectiveness of screening prioritization by means of active learning models [@Yu2018; @Yu2019; @Miwa2014; @Cormack2014; @Cormack2015; @Wallace2010; @Gates2018a]. With active learning, the ML model can iteratively improve its predictions on unlabelled data by allowing the model to select the records from which it wants to learn [@Settles2012]. The model queries these records to a human annotator who provides them with a label, from which the model then updates its predictions. The general assumption is that by letting the model select which records are labelled, the model can achieve higher accuracy more quickly while requiring the human annotator to label as few records as possible [@Settles2009]. Active learning has proven to be an efficient strategy in large unlabelled datasets where labels are expensive to obtain [@Settles2009]. This makes the screening phase in systematic reviewing an ideal candidate solution for such models because typically, labelling a large number of publications is very costly. When active learning is applied in the screening phase, the reviewer screens publications that are suggested by an active learning model. Subsequently, the active learning model learns from the reviewers' decision ('relevant', 'irrelevant') and uses this knowledge to update its predictions and to select the next publication to be screened by the reviewer.

<!-- Combined with some sort of stopping criterion, the reviewer c -->
<!-- _Adopting some sort of stopping criterion (outside scope of the current thesis) the reviewer can quit reviewing after having read only a fraction of candidate studies. meaning the screening process can be finished after reading a fraction of all candidate studies. Saving hours of time and resources._ -->

@@

The application of active learning models in reducing workload of systematic reviews has been extensively studied [@Yu2018; @Yu2019; @Miwa2014; @Wallace2010; @Gates2018a]. Whilst previous studies have evaluated active learning models in many forms and shapes [@Yu2018; @Yu2019; @Miwa2014; @Cormack2014; @Cormack2015; @Wallace2010; @Singh2018; @Carvallo; @Yimin2007], ready-to-use software tools implementing such active learning models (Abstrackr [@Wallace2012b], Colandr [@Cheng2018], FASTREAD [@Yu2018], Rayyan [@Ouzzani2016], and RobotAnalyst [@Przybyla2018]) currently use the same classification technique to predict relevance of publications, namely Support Vector Machine (SVM). Findings show that different classification techniques can serve different needs in the retrieval of relevant publications, for example the desired balance between recall and precision [@Kilicoglu2009; @Aphinyanaphongs2004]. Therefore, is essential to evaluate different classification techniques in the context of active learning models. The current study investigates active learning models adopting four classification techniques: Naive Bayes, Logistic Regression, Support Vector Machine, and Random Forest. These are widely adopted techniques in text classification [@Aggarwal2012] and are suitable for creating interpretable ML models [@Molnar2020]. Moreover, due to their relatively short computation time they are fit for software tools to be used in scientific practice, as they allow the reviewer to be timely presented with the next publication.

@@

<!-- no study in this area has made a comparison between models adopting different classification techniques. -->
<!-- Since a plethora of techniques exist [@Aggarwal2012], it would be of interest to incorporate such models to this area as well.  -->
Another component that influences model performance is the way how the textual content of titles and abstracts are represented in a model, called the feature extraction strategy [@Zhang2011; @Carvallo; @Singh2018]. One of the more sophisticated feature extraction strategies is called Doc2vec (D2V), also known as Paragraph vector [@Le2014]. D2V learns continuous distributed vector representations for pieces of text. The idea of distributed representation is that every word can be known by its neighborhood. D2V can overcome problems with more simplistic feature extraction strategies such as TF-IDF, which ignores the the semantics and the ordering of the words [@Le2014]. Therefore, it is of interest to compare models adopting D2V to models adopting TF-IDF.

@@

Lastly, previous studies have mainly focussed on reviews from a single scientific field, like medicine [@Wallace2010; @Gates2018a] and computer science [@Yu2018, @Yu2019]. Model replications on reviews from varying research contexts are essential to draw conclusions about the general effectiveness of active learning models [@OMara-Eves2015; @Marshall2020]. As far as known to the authors, Miwa et al [@Miwa2014] were the only researchers to make a direct comparison between two systematic reviews from different research areas, namely the social and the medical sciences. They found that active learning was more difficult on data from the social sciences due to the different nature of the vocabularies used. Thus, it is of interest to evaluate model performance across different research contexts, namely the social science, humanities, medical science and computer science.

@@

|    __RQ1__ What is the performance of active learning models across four different classification techniques?
|    __RQ2__ What is the performance of active learning models across two different feature extraction strategies? 
|    __RQ3__ Does the performance of active learning models differ across six systematic reviews from four different research areas?

The purpose of the current paper is to increase the evidence base on active learning models for reducing workload in title and abstract screening in systematic reviews. We adopt four different classification techniques (NB, LR, SVM, and RF) and two different feature extraction strategies (TF-IDF and D2V) for the purpose of maximizing the number of identified relevant publications, while minimizing the number of publications needed to screen. Models were assessed by conducting a simulation on six systematic review datasets. Datasets were collected from the fields of medical science [@Cohen2006; @Appenzeller-Herzog2019; @Kwok2020], computer science [@Yu2018], humanities [@Nagtegaal2019] and social science [@vandeSchoot2017], to assess generalizability of the models across research contexts. The models, datasets and simulations are implemented in a pipeline of active learning for screening prioritization, called $\texttt{ASReview}$ [@vandeSchoot2020]. $\texttt{ASReview}$ is an open source and generic tool in which users can adapt and add modules as they like, encouraging fellow researchers to replicate findings from previous studies. All scripts and data used are openly published to facilitate usability and acceptability of ML-assisted title and abstract screening in the field of systematic review.

The remaining part of this paper is organized as follows. The methods section describes the study that was designed to answer the research questions and elaborates on the technical details of active learning models for identifying relevant publications in the context of systematic reviews. The findings of the simulation study are reported in the results section. The implications of the findings in context of previous research are discussed in the discussion section, followed by this study's main conclusions in the conclusion section.

@@

## Technical details 
What follows is a more detailed account of the active learning models. The structure and functions of the key components of the models are introduced to clarify the choices made in the design of the current study.

### Task description
The screening process of a systematic review starts with all publications obtained in the search. The task is to identify which of these publications are relevant, by screening their titles and abstracts. In active learning for screening prioritization, the screening process proceeds as follows:

<!-- Just a sample algorithmn -->
<!-- \begin{algorithm}[H] -->
<!-- \DontPrintSemicolon -->
<!-- \SetAlgoLined -->
<!-- \KwResult{Write here the result} -->
<!-- \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output} -->
<!-- \Input{Write here the input} -->
<!-- \Output{Write here the output} -->
<!-- \BlankLine -->
<!-- \While{While condition}{ -->
<!--     instructions\; -->
<!--     \eIf{condition}{ -->
<!--         instructions1\; -->
<!--         instructions2\; -->
<!--     }{ -->
<!--         instructions3\; -->
<!--     } -->
<!-- } -->
<!-- \caption{While loop with If/Else condition} -->
<!-- \end{algorithm} -->

- Start with the set of all unlabelled records (titles and abstracts), $\mathcal{U}$.
- The reviewer provides a label for a few, for example 5-10 records $x \in \mathcal{U}$, creating a set of labelled records $x \in \mathcal{L}$ such that $x \notin \mathcal{U}$. The label can be either Relevant $\langle x, \texttt{R}\rangle$ or Irrelevant $\langle x, \texttt{I}\rangle$.
- The active learning cycle starts:

  1. A classifier $C$ is trained on the labelled records $\mathcal{L}$, $C = \texttt{train}(\mathcal{L})$

  2. The classifier predicts relevancy scores for all unlabelled records $\mathcal{U}$, $C(\mathcal{U})$

@@

  4. The model queries the reviewer to screen this record, $\langle x^*, \texttt{?}\rangle$

@@

In this active learning cycle, the model incrementally improves its predictions on the remaining unlabelled title and abstracts. Relevant titles and abstracts are identified as early in the process as possible.

This case is an example of pool-based active learning, as the next record to be queried is selected by predicting relevancy for all records in a fixed pool [@Settles2012]. Another form of active learning is stream-based active learning, in which data is regarded as a stream instead of a fixed pool, in which the model selects one record at a time and then decides whether or not to query this record. This approach of active learning is preferred when it is expensive or impossible to exhaustively search the data for selecting the next query. A possible application of stream-based active learning is living systematic reviews, as the review is continually updated as new evidence becomes available. For an example see the study by Wynants et al., @Wynants2020.

### Class imbalance problem
There are two classes in the dataset: relevant and irrelevant publications. Typically, the inclusion rate is low as only a fraction of the publications belong to the relevant class (2.94%, [@Borah2017]). The class imbalance causes the classifier to miss relevant publications, because there are far fewer examples of relevant than irrelevant publications to train on [@OMara-Eves2015]. Moreover, classifiers can achieve high accuracy but still fail to identify any of the relevant publications [@Wallace2010]. This is evident in the case of a systematic review dataset where only three percent of publications are relevant.

@@

Previous studies have addressed the class imbalance problem by rebalancing the training data in various ways [@OMara-Eves2015]. To decrease the class imbalance in the training data, the models in the current study rebalance the training set by a technique we propose to call Dynamic Resampling (DR). DR undersamples the number of irrelevant publications in the training data, whereas the number of relevant publications are oversampled such that the size of the training data remains the same. The ratio between relevant and irrelevant publications in the rebalanced training data is not fixed, but dynamically updated depending on the number of publications in the available training data, the total number of publications in the dataset, and the ratio between relevant and irrelevant publications in the available training data.

@@

### Classification
To make predictions on the unlabelled publications, a classifier is trained on features from the training data. The performance of the following four classifiers is explored:
<!-- The classifier predicts class of a given unlabelled publication.  -->
<!-- Whilst the SVM is a well-known technique, also classifiers have been employed in class prediction tasks in systematic reviews.  -->
<!-- In fact, research in this area to date has not yet determined the performance of classifiers other than SVMs. -->

 - Support Vector Machine (SVM) - SVMs separate the data into classes by finding a multidimensional hyperplane [@Tong2001; @Kremer2014].

 - L2-regularized Logistic Regression (LR) models the probabilities describing the possible outcomes by a logistic function. The coefficients are regularizes/model is regularized. 
 A penalty is imposed on the size of the coefficients, shrinking coefficients of features with a minor contribution to the solution towards zero.

@@

- Random Forests (RF) is a supervised learning algorithm where a large number of decision trees are fit on bootstrapped samples of the original data. All trees cast a vote on the class, which are aggregated into a class prediction for each instance [@Breiman2001].

### Feature extraction
To predict relevance of a given publication, the classifier uses information from the publications in the dataset. Examples of such information are titles and abstracts. However, a model cannot make predictions from the titles and abstracts as they are; their textual content needs to be represented numerically. The textual information needs to be mapped to feature vectors. This process of numerically representing textual content is referred to as 'feature extraction'.

Term-Frequency Inverse Document Frequency (TF-IDF) stores the term frequency - the number of occurrences of each word - for each text in the dataset. The term frequency is adjusted with the inverse document frequency, the number of documents which contain the given word [@Ramos2003]. A disadvantage of TF-IDF and other bag-of-words methods is that they do not take the ordering of words into account, thereby ignoring semantics. An example of an approach that aims to overcome this weakness is D2V. D2V learns continuous distributed vector representations for pieces of text capable of grasping semantics by learning to predict the words in the texts [@Le2014].

>>>>

---
title: Active learning for screening prioritization in systematic reviews
subtitle: A simulation study
author: 
    - Gerbrich Ferdinands
    - Raoul Schram
    - Jonathan de Bruin
    - Ayoub Bagheri
    - Daniel Oberski
    - Lars Tummers
    - Rens van de Schoot
date: '`r format(Sys.time(), "%d %B, %Y")`'
header-includes:
    - \usepackage{setspace}\doublespacing

output:
  bookdown::pdf_document2:
    extra_dependencies: subfig, algorithm2e, float, lineno, setspace, amssymb
    number_sections: no
    toc: no
  word_document: default
bibliography: asreview.bib
link-citations: yes
csl: systematic-reviews.csl
---

<!-- --- -->
<!-- title: Interactive screening prioritization in systematic reviews by employing active learning -->
<!-- subtitle: A comprehensive study -->
<!-- author:  -->
<!--     - Gerbrich Ferdinands -->
<!--     - Raoul Schram -->
<!--     - Jonathan de Bruin -->
<!--     - Ayoub Bagheri -->
<!--     - Daniel Oberski -->
<!--     - Lars Tummers -->
<!--     - Rens van de Schoot -->
<!-- date: '`r format(Sys.time(), "%d %B, %Y")`' -->
<!-- header-includes: -->
<!--     - \usepackage{setspace}\doublespacing -->
<!-- output: -->
<!--   word_document: default -->
<!--   bookdown::pdf_document2: -->
<!--     extra_dependencies: subfig, algorithm2e, float, lineno, setspace, amssymb -->
<!--     number_sections: no -->
<!--     toc: no -->
<!-- bibliography: asreview.bib -->
<!-- link-citations: yes -->
<!-- csl: systematic-reviews.csl -->
<!-- --- -->

@@

```{r all tables, results = 'hide'}
table_nums <- captioner(prefix = "Table")
# table  on datasets statistics
table_nums(name = "datasets", caption = "Statistics on the datasets obtained from six original systematic reviews.")
table_nums(name = "atd", caption = "ATD values ($\\bar x (\\hat s)$) for all model-dataset combinations. For every dataset, the best results are in bold. Median (MAD) is given for all datasets.")
table_nums(name = "wss", caption = "WSS@95 values ($\\bar x (\\hat s)$) for all model-dataset combinations. For every dataset, the best results are in bold. Median (MAD) is given for all datasets.")
table_nums(name = "rrf", caption = "RRF@10 values ($\\bar x, (\\hat s)$) for all model-dataset combinations. For every dataset, the best results are in bold. Median (MAD) is given for all datasets.")

@@

\linenumbers
# Abstract
#### Background
Conducting a systematic review requires great screening effort. Various tools have been proposed to speed up the process of screening thousands of titles and abstracts by engaging in active learning. In active learning, the reviewer interacts with machine learning software to identify relevant publications as early as possible. To gain a comprehensive understanding of active learning models for reducing workload in systematic reviews, the current study provides an methodical overview of such models. Active learning models were evaluated across four different classification techniques (naive Bayes, logistic regression, support vector machines, and random forest) and two different feature extraction strategies (TF-IDF and Doc2vec). Moreover, models were evaluated across six systematic review datasets from various research areas to assess generalizability of active learning models across different research contexts.

#### Methods
Performance of the models were assessed by conducting simulations on six systematic review datasets. We defined desirable model performance as maximizing recall while minimizing the number of publications needed to screen. Model performance was evaluated by recall curves, WSS@95, RRF@10, and ATD.

#### Results
Within all datasets, the model performance exceeded screening at random order to a great degree, reducing the number of publications needed to screen by 91.7% to 63.9%. Overall, the Naive Bayes + TF-IDF model performed the best.

#### Conclusions
Active learning models for screening prioritization show great potential in reducing the workload of systematic reviewers.
<!-- Model performance differs vastly across datasets, and the naive Bayes classifier is a  -->
<!-- The results shed new light on the performance of different classification techniques, indicating that the NB classification technique is superior to the widely used SVM. As model performance differs vastly across datasets,  -->
<!-- To facilitate the usability of active learning models in systematic review practice, it is essential to identify how dataset characteristics relate to model performance. -->

__keywords:__ systematic reviews, active learning, screening prioritization, researcher-in-the-loop, title-and-abstract screening

<!-- #### Background  -->
<!-- the context and purpose of the study -->

<!-- context: -->
<!-- active learning models for reducing workload in title and abstract screening in systematic reviews. -->

<!-- This study adds to existing literature by evaluating models across four different classification techniques (Naive Bayes, Logistic Regression, Support Vector Machine, and Random Forest) and two different feature extraction strategies (TF-IDF and Doc2vec). Moreover, as previous studies have paid little attention to the generalizability of active learning models across different research contexts, models were evaluated across six systematic review datasets from various research areas. -->
<!-- purpose  -->

<!-- The purpose of the current paper is to increase the evidence base on active learning models for reducing workload in title and abstract screening in systematic reviews.  -->

<!-- for the purpose of maximizing the number of identified relevant publications, while minimizing the number of publications needed to screen.  -->

<!--  We adopt four different classification techniques (Naive Bayes, Logistic Regression, Support Vector Machine, and Random Forest) and two different feature extraction strategies (TF-IDF and Doc2vec) for the purpose of maximizing the number of identified relevant publications, while minimizing the number of publications needed to screen.  -->

<!-- #### Methods -->
<!-- how the study was performed and statistical tests used -->

<!-- Models were assessed by conducting a simulation on six systematic review datasets. -->
<!-- To assess generalizability of the models across research contexts, datasets were collected from the fields of medical science, computer science, humanities and social science,  -->

<!-- Desirable model performance was defined as maximizing recall while minimizing the number of publications needed to screen. (WSS@95 recall curves RRF@10 ATD) -->

<!-- #### Results -->
<!-- the main findings -->

<!-- Across all classification techniques, feature extraction strategies and review contexts, active learning models exceeded screening at random order by far.  -->

<!-- #### Conclusions -->
<!-- brief summary and potential implications -->

<!-- As all models exceeded screening at random order by far, they show great potential in reducing workload of systematic reviewers.  -->

<!-- implications: researchers can use this  -->

@@

# Background
Systematic reviews are top of the bill in research. A systematic review brings together all available studies relevant to answer a specific research question [@PRISMA-PGroup2015]. Systematic reviews inform practice and policy [@Gough2002] and are key in developing clinical guidelines [@Chalmers2007]. However, systematic reviews are costly because to identify publications relevant to answering the research question, they among else involve the manual screening of thousands of titles and abstracts.

<!-- An experienced reviewer takes on average 30 seconds to screen one title and abstract, whereas an inexperienced reviewer takes even longer [@Wallace2010]. A typical review of 5000 publications would require over 40 hours of uninterrupted screening, while typically only a fraction of the publications is relevant (mean = 2.94%, IQR=2.5 @Borah2017]). -->
Conducting a systematic review typically requires over a year of work by a team of researchers [@Borah2017]. Nevertheless, systematic reviewers are often bound to a limited budget and timeframe. Currently, the demand for systematic reviews exceeds the available time and resources by far [@Lau2019]. Especially when answering the research question at hand is urgent, it is extremely challenging to provide a review that is both timely and comprehensive.

To ensure a timely review, reducing the workload in systematic reviews is essential. With advances in machine learning (ML), there has been wide interest in tools to reduce the workload in systematic reviews [@Harrison2020]. Various ML models have been proposed, aiming to predict whether a given publication is relevant or irrelevant to the systematic review. Previous findings suggest that such models potentially reduce the workload with 30-70% at the cost of losing 5% of relevant publications, in else, a 95% recall [@OMara-Eves2015].
<!-- To date, several studies have begun to examine the use of learning algorithms for screening prioritization. -->

A well-established approach to increase the efficiency of title and abstract screening is screening prioritization [@Cohen2009; @Shemilt2014]. In screening prioritization, the ML model presents the reviewer with the publications that are most likely to be relevant first, thereby expediting the process of finding all of the relevant publications. Such an approach allows for substantial time-savings in the screening process as the reviewer can decide to stop screening after a sufficient number of relevant publications have been found [@Yu2019]. Moreover, the early retrieval of relevant publications facilitates a faster transition of those publications to the next steps in the review process [@Cohen2009]. 
<!-- Additionally, several studies report increasing efficiency beyond saving time [@OMara-Eves2015].  -->

<!-- assist the reviewer in the screening process -->
<!-- Several studies have proposed active learning models for screening prioritization -->
Recent studies have demonstrated the effectiveness of screening prioritization by means of active learning models [@Yu2018; @Yu2019; @Miwa2014; @Cormack2014; @Cormack2015; @Wallace2010; @Gates2018a]. With active learning, the ML model can iteratively improve its predictions on unlabeled data by allowing the model to select the records from which it wants to learn [@Settles2012]. The model proposes these records to a human annotator who provides the records with labels, which the model then uses to update its predictions. The general assumption is that by letting the model select which records are labelled, the model can achieve higher accuracy more quickly while requiring the human annotator to label as few records as possible [@Settles2009]. Active learning has proven to be an efficient strategy in large unlabeled datasets where labels are expensive to obtain [@Settles2009]. This makes the screening phase in systematic reviewing an ideal candidate solution for such models, because typically labelling a large number of publications is very costly.

When active learning is applied in the screening phase, the reviewer screens publications that are suggested by an active learning model. Subsequently, the active learning model learns from the reviewers' decision ('relevant', 'irrelevant') and uses this knowledge to update its predictions and to select the next publication to be screened by the reviewer.

<!-- Combined with some sort of stopping criterion, the reviewer c -->
<!-- _Adopting some sort of stopping criterion (outside scope of the current thesis) the reviewer can quit reviewing after having read only a fraction of candidate publications. meaning the screening process can be finished after reading a fraction of all candidate publications. Saving hours of time and resources._ -->

@@

The application of active learning models in systematic reviews has been extensively studied [@Yu2018; @Yu2019; @Miwa2014; @Wallace2010; @Gates2018a]. While previous studies have evaluated active learning models in many forms and shapes [@Yu2018; @Yu2019; @Miwa2014; @Cormack2014; @Cormack2015; @Wallace2010; @Singh2018; @Carvallo; @Yimin2007], ready-to-use software tools implementing such models (Abstrackr [@Wallace2012b], Colandr [@Cheng2018], FASTREAD [@Yu2018], Rayyan [@Ouzzani2016], and RobotAnalyst [@Przybyla2018]) currently use the same classification technique to predict relevance of publications, namely support vector machines (SVM). It was found [@Kilicoglu2009; @Aphinyanaphongs2004] that different classification techniques can serve different needs in the retrieval of relevant publications, for example the desired balance between recall and precision. Therefore, it is essential to evaluate different classification techniques in the context of active learning models. The current study investigates active learning models adopting four classification techniques: naive Bayes (NB), logistic regression (LR), SVM, and random forest (RF). These are widely adopted techniques in text classification [@Aggarwal2012] and are suitable for creating interpretable ML models [@Molnar2020]. Moreover, due to their relatively short computation time they are fit for software tools to be used in scientific practice.

<!-- as they allow the reviewer to be timely presented with the next publication.  -->

@@

<!-- no study in this area has made a comparison between models adopting different classification techniques. -->
<!-- Since a plethora of techniques exist [@Aggarwal2012], it would be of interest to incorporate such models to this area as well.  -->
Another component that influences model performance is how the textual content of titles and abstracts is represented in a model, called the feature extraction strategy [@Zhang2011; @Carvallo; @Singh2018]. One of the more sophisticated feature extraction strategies is Doc2vec (D2V), also known as paragraph vectors [@Le2014]. D2V learns continuous distributed vector representations for pieces of text. In distributed text representations, words are assumed to appear in the same context when they are similar in terms of a latent space, the “embedding”. A word embedding is simply a vector of scores estimated from a corpus for each word; D2V is an extension of this idea to document embeddings. Embeddings can sometimes outperform simpler feature extraction strategies such as term frequency-inverse document frequency (TF-IDF). They can be trained on large corpora to capture wider semantics and subsequently applied in a specific systematic reviewing application [@Le2014]. Therefore, it is interesting to compare models adopting D2V to models adopting TF-IDF.

@@

Lastly, previous studies have mainly focussed on reviews from a single scientific field, like medicine [@Wallace2010; @Gates2018a] or computer science [@Yu2018, @Yu2019]. To draw conclusions about the general effectiveness of active learning models, it is essential to evaluate models on reviews from varying research contexts [@OMara-Eves2015; @Marshall2020]. To our knowledge, Miwa et al [@Miwa2014] were the only researchers to make a direct comparison between systematic reviews from different research areas, such as the social and the medical sciences. They found that the application of active learning to systematic reviews was more difficult on a systematic review from the social sciences due to the different nature of the vocabularies used. Thus, it is of interest to evaluate model performance across different research contexts, namely social science, medical science and computer science.

@@

__RQ1__ What is the performance of active learning models across four classification techniques?
__RQ2__ What is the performance of active learning models across two feature extraction strategies? 
__RQ3__ Does the performance of active learning models differ across six systematic reviews from four research areas?

The purpose of this paper is to show the usefulness of active learning models for reducing workload in title and abstract screening in systematic reviews. We adopt four different classification techniques (NB, LR, SVM, and RF) and two different feature extraction strategies (TF-IDF and D2V) for the purpose of maximizing the number of identified relevant publications, while minimizing the number of publications needed to screen. Models were assessed by conducting a simulation on six systematic review datasets. To assess generalizability of the models across research contexts, datasets containing previous systeamtic reviews were collected from the fields of medical science [@Cohen2006; @Appenzeller-Herzog2019; @Kwok2020], computer science [@Yu2018], and social science [@Nagtegaal2019; @vandeSchoot2017]. The models, datasets and simulations are implemented in a pipeline of active learning for screening prioritization, called $\texttt{ASReview}$ [@vandeSchoot2020]. $\texttt{ASReview}$ is an open source and generic tool in which users can adapt and add modules as they like, encouraging fellow researchers to replicate findings from previous studies. To facilitate usability and acceptability of ML-assisted title and abstract screening in the field of systematic review our scripts and data used are openly available.

<!-- The remaining part of this paper is organized as follows. The methods section describes the study that was designed to answer the research questions and elaborates on the technical details of active learning models for identifying relevant publications in the context of systematic reviews. The findings of the simulation study are reported in the results section. The implications of the findings in context of previous research are discussed in the discussion section, followed by this study's main conclusions in the conclusion section.  -->

@@

## Technical details 
<!-- What follows is a more detailed account of the active learning models. The structure and functions of the key components of the models are introduced to clarify the choices made in the design of the current study.   -->
What follows is a more detailed account of the active learning models to clarify the choices made in the design of the current study.

### Task description
The screening process of a systematic review starts with all publications obtained in the search. The task is to identify which of these publications are relevant, by screening their titles and abstracts. In _active learning_ for screening prioritization, the screening process proceeds as follows:

<!-- Just a sample algorithmn -->
<!-- \begin{algorithm}[H] -->
<!-- \DontPrintSemicolon -->
<!-- \SetAlgoLined -->
<!-- \KwResult{Write here the result} -->
<!-- \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output} -->
<!-- \Input{Record IDs in screening order} -->
<!-- \Output{Average Time to Discovery} -->
<!-- \BlankLine -->
<!--  # inclusions -->
<!--  one_labels = np.where(labels == 1)[0] -->
<!--  # store discovery time -->
<!--  time_results = {label: [] for label in one_labels} -->

<!-- \While{While condition}{ -->
<!--     instructions\; -->
<!--     \eIf{condition}{ -->
<!--         instructions1\; -->
<!--         instructions2\; -->
<!--     }{ -->
<!--         instructions3\; -->
<!--     } -->
<!-- } -->
<!-- \caption{While loop with If/Else condition} -->
<!-- \end{algorithm} -->

- Start with the set of all unlabeled records (titles and abstracts), $\mathcal{U}$.
- The reviewer provides a label for a few, e.g. 5-10 records $x \in \mathcal{U}$, creating a set of labelled records $x \in \mathcal{L}$ such that $x \notin \mathcal{U}$. The label can be either Relevant $\langle x, \texttt{R}\rangle$ or Irrelevant $\langle x, \texttt{I}\rangle$.
- The active learning cycle starts:

  1. A classifier $C$ is trained on the labelled records $\mathcal{L}$; $C = \texttt{train}(\mathcal{L})$

  2. The classifier predicts relevancy scores for all unlabeled records $\mathcal{U}$, $C(\mathcal{U})$

@@

  4. The model requests the reviewer to screen this record, $\langle x^*, \texttt{?}\rangle$

@@

In this active learning cycle, the model incrementally improves its predictions on the remaining unlabeled title and abstracts. Relevant titles and abstracts are identified as early in the process as possible.

This case is an example of pool-based active learning, as the next record to be queried is selected by predicting relevancy for all records in a fixed pool [@Settles2012]. Another form of active learning is stream-based active learning, in which the data is regarded as a stream instead of a fixed pool, in which the model selects one record at a time and then decides whether or not to query this record. This approach of active learning is preferred when it is expensive or impossible to exhaustively search the data for selecting the next query. A possible application of stream-based active learning is living systematic reviews, as the review is continually updated as new evidence becomes available. For an example see the study by Wynants et al @Wynants2020.

### Class imbalance problem
<!-- There are two classes in the dataset: relevant and irrelevant publications.  -->
Typically, only a fraction of the publications belong to the relevant class (2.94%, [@Borah2017]). To some extent, this fraction is under the control of the researcher through the search criteria: if the researcher narrows the search query, it will generally result in a higher proportion of relevant publications. However, in most applications this practice would yield an unacceptable number of false negatives (erroneously excluded papers) in the querying phase of the review process. For this reason, the query phase in most practical applications would typically yield a very low percentage of relevant publications. In the discussion, we will provide some further remarks on the interplay among query scope and subsequent active machine learning in the abstract screening phase. Because there are genearlly far fewer examples of relevant than irrelevant publications to train on, the class imbalance causes the classifier to miss relevant publications [@OMara-Eves2015]. Moreover, classifiers can achieve high accuracy but still fail to identify any of the relevant publications [@Wallace2010].

<!-- This is evident in the case of a systematic review dataset where only three percent of publications are relevant.  -->

@@

Previous studies have addressed the class imbalance problem by rebalancing the training data in various ways [@OMara-Eves2015]. To decrease the class imbalance in the training data, we rebalance the training set by a technique we propose to call "dynamic resampling" (DR). DR undersamples the number of irrelevant publications in the training data, whereas the number of relevant publications are oversampled such that the size of the training data remains the same. The ratio between relevant and irrelevant publications in the rebalanced training data is not fixed, but dynamically updated and depends on the number of publications in the available training data, the total number of publications in the dataset, and the ratio between relevant and irrelevant publications in the available training data. Additional file 1 provides a detailed script to perform DR.

@@

### Classification
To make relevancy predictions on the unlabeled publications, a classifier is trained on features from the training data. The performance of the following four classifiers is explored:
<!-- The classifier predicts class of a given unlabeled publication.  -->
<!-- Whilst the SVM is a well-known technique, also classifiers have been employed in class prediction tasks in systematic reviews.  -->
<!-- In fact, research in this area to date has not yet determined the performance of classifiers other than SVMs. -->

 - Support vector machines (SVM) - SVMs separate the data into classes by finding a multidimensional hyperplane [@Tong2001; @Kremer2014].

 - L2-regularized logistic regression (LR) - models the probabilities describing the possible outcomes by a logistic function. The classifier uses regularization, shrinking coefficients of features with small contributions to the solution towards zero.

<!-- A penalty is imposed on the size of the coefficients, shrinking coefficients of features with a minor contribution to the solution towards zero. -->

@@

- Random forests (RF) is a supervised learning algorithm where a large number of decision trees are fit on samples obtained from the original data by sampling both rows (bootstrapped samples) and columns (feature samples). In prediction mode, each tree casts a vote on the class, and the final prediction is the class that received the most votes [@Breiman2001].

### Feature extraction
To predict relevance of a given publication, the classifier uses information from the publications in the dataset. Examples of information are titles and abstracts. However, a model cannot make predictions from the titles and abstracts as they are; their textual content needs to be represented numerically as feature vectors. This process of numerically representing textual content is referred to as 'feature extraction'.
